Centers and centralizers of higher enveloping algebras of Lie algebras

### 2020-09-09 draft.tex
moved and improved the cyclic stuff

Recently, a generalization of the concept of the enveloping algebra of a Lie algebra has emerged: 
each dg Lie algebra $\fg$ has an enveloping $\EE_n$ algebra $\UU_n\fg$, for every $n$, 
so that $\UU_1\fg$ is the usual enveloping algebra $U\fg$. 
Building on prior work \cite{BD, AF} [Owen's thesis?], 
Knudsen \cite{Knudsen} developed these ideas extensively and offered a useful factorization algebra model for the enveloping $\EE_n$ algebra. 
Our central goal here is to generalize the some well-known results about the enveloping algebra $U\fg$ to $\UU_n\fg$, for all $n$, such as
\begin{enumerate}
\item[(1)] the (underived) center of $U\sl_{N+1}$ is isomorphic to $(\Sym(\CC^{N}))^{S_N}$, where $\CC^{N}$ denotes the Cartan subalgebra of $\sl_{N+1}$ and $S_N$ is the Weyl group of $SL_{N+1}$, and
\item[(2)] the derived center of the enveloping algebra $U\fg$ is given by derived invariants of the adjoint action of $\fg$ on $U\fg$. That is to say $\Hoch^*(U\fg,U\fg) \simeq \clies(\fg, U\fg)$.
\end{enumerate}
\owen{And keep going just a little.}
%
%Our results pertains to (1) by taking $N \to \infty$, and to (2) through the derived {\em centralizer} of the map $\UU_n\sl_\infty \hookrightarrow \UU_n\gl_\infty(A)$, where $A$ is a dg algebra. Note that the usual derived center is the centralizer of the identity map $\UU_n\fg \to \UU_n\fg$.
%
%We will review these notions and describe our results in detail Min the next section. When $n=1$,  \owen{state the classical thing here!}
%
%Our primary tool here is the Loday-Quillen-Tsygan (LQT) theorem which describes the large $N$ limit of the Lie algebra homology of $\gl_N(A)$, for a dg associative algebra $A$. Since the enveloping $\EE_n$ algebra of a Lie algebra can be constructed as a locally constant factorization algebra made out of the shifted Chevalley-Eilenberg chains \cite{}, the relevance of the LQT theorem in this context  should be apparent. From this point of view, one can think of our main result as an $\EE_n$ generalization of the LQT theorem. We hope these ideas have exciting applications in quantum field theory and random matrix theory, as well as to other large $N$ limit phenomena.

@@

We apply our central result to important classes of Lie algebras: 
abelian, semisimple, and free.
We also indicate the large $N$ behavior of enveloping algebras for the infinite families $\frak{gl}_N$, $\frak{sp}_N$, and $\frak{so}_N$,
by exploiting the Loday-Quillen-Tsygan theorem and its cousins.

@@

\subsection{The large $N$ limit of the center ${\frak Z}_n(\UU_n \frak{gl}_N)$ and cyclic cohomology}

\owen{Please check this carefully, Greg. The lemma is meaningful even if the proposition is bogus.}

In the case of the infinite series of Lie algebras, 
there are theorems that tell us about the limiting behavior of Lie algebra chains and cochains as $N$ goes to infinity.
These can be applied to our models of the center to obtain novel $\EE_n$ algebras.

Let $\Cyc_*(A)$ denote the {\em cyclic chains} of a dg associative algebra~$A$.

\def\gl{\frak{gl}}

\begin{prp}
There is a quasi-isomorphism of $\EE_n$ algebras
\[
{\frak Z}_n(\frak{gl}_\infty) \xto{\simeq} \clies(\frak{\gl}_\infty) \otimes \Sym(\Cyc_*(k[-n])[1])  ,
\]
and so, in particular, the $N \to \infty$ limit of ${\frak Z}_n(\frak{gl}_N)$ is commutative.
\end{prp}

\begin{proof}
The Loday-Quillen-Tsygan (LQT) theorem --- for nonunital algebras --- provides a quasi-isomorphism
\[
\cliels(\gl_\infty(A))^{\gl_\infty(k)} \xto{\simeq} \Sym(\Cyc_*(A)[1])
\] 
and it is functorial in the algebra~$A$.
\owen{I used {\em invariants} here rather than {\em coinvariants}. Is that allowed?}
Now
\[
\clies(\gl_\infty, \cliels(\gl_\infty(A)) \simeq \cliels(\gl_\infty) \otimes \cliels(\gl_\infty(A))^{\gl_\infty(k)},
\]
as shown in the proof of the LQT theorem.

We take $A = \Omega^*_c(V)$ and use the functoriality to see that we get a quasi-isomorphism of prefactorization algebras, as claimed.
\end{proof}

There are immediate analogs of this theorem where we replace $\gl_N$ by $\frak{so}_N$ or $\frak{sp}_N$ and where we replace cyclic chains by dihedral chains.

A striking consequence of the LQT map appears at the level of factorization homology.

\begin{lmm}
For a closed $n$-dimensional manifold~$M$, 
\begin{align*}
\int_M {\frak Z}_n(\frak{gl}_\infty) 
&\simeq \clies(\Omega^*(M) \otimes \gl_\infty, \Sym(\Cyc_*(\Omega^*(M))[1])) \\
&\simeq \clies(\Omega^*(M) \otimes \gl_\infty, \Sym(\Omega^*(LM/^h S^1))[1]))
\end{align*}
where $\Omega^*(LM/^h S^1)$ denotes some cochain model for the cochains of the homotopy quotient of the free loop space under the canonical rotation action on loops.
\end{lmm}

\owen{That phrasing could be improved.}

Implicit in the statement is that one must describe the needed (and non-obvious) action of $\gl_\infty(\Omega^*(M))$ on a symmetric algebra generated by cyclic chains, respectively the rotation-equivariant cochains on the free loop space~$LM$.

What makes this result striking is that it shows how the {\em cyclic loop space} $LM/^h S^1$ appears out of taking a large $N$ limit of factorization homology.
This kind of result is especially interesting from the perspective of physics, 
as a kind of topological gauge/string duality.
In Section~\ref{sec: TFT} we explain how the center $\frak{Z}_n(\gl_\infty)$ can be understood as the {\em quantum observables} of topological BF theory for the Lie algebra $\gl_\infty$.
In this sense, the left hand side of the lemma $\int_M {\frak Z}_n(\frak{gl}_\infty)$ describes the global quantum observables over the manifold $M$.
On the right hand side, we see the cyclic loop space of $M$,
which arises naturally as the configuration space of a closed string theory with target~$M$.

\begin{proof}
As $\Omega^*_c(M) = \Omega^*(M)$ because $M$ is closed, 
one applies the LQT theorem for {\em unital} algebras
\[
\cliels(\gl_\infty(\Omega^*(M))) \xto{\simeq} \Sym(\Cyc_*(\Omega^*(M))[1])
\]
to obtain the first weak equivalence.
(In fact, one must transfer the $\gl_\infty(\Omega^*(M))$ along the LQT map.)
Then Jones' theorem \cite{} provides the second weak equivalence.
\end{proof}

@@

In the case $k = 0$, we have the immediate corollary that
\[
\int_{S} \fZ_n(\fg) \simeq \fZ_0 \left( \int_S \UU_n \fg \right) ,
\]
which unwinds to 
\[
\clies(\fg \otimes \Omega^\bu(S), \cliels(\fg \otimes \Omega^\bu(S)) \simeq \End^\bullet(\cliels(\fg \otimes \Omega^\bu(S))),
\]
which looks rather tautological.
Nonetheless, by using other models for factorization homology,
one obtains less obvious identifications.

@@

>>>>

Recently, a generalization of the concept of the enveloping algebra of a Lie algebra has emerged: 
each dg Lie algebra $\fg$ has an enveloping $\EE_n$ algebra $\UU_n\fg$, for every $n$, 
so that $\UU_1\fg$ is the usual enveloping algebra $U\fg$. 
Building on prior work \cite{BD, AF} [Owen's thesis?], 
Knudsen \cite{Knudsen} developed these ideas extensively and offered a useful factorization algebra model for the enveloping $\EE_n$ algebra. 
Our central goal here is to generalize the some well-known results about the enveloping algebra $U\fg$ to $\UU_n\fg$, for all $n$, such as
\begin{enumerate}
\item[(1)] the (underived) center of $U\sl_{N+1}$ is isomorphic to $(\Sym(\CC^{N}))^{S_N}$, where $\CC^{N}$ denotes the Cartan subalgebra of $\sl_{N+1}$ and $S_N$ is the Weyl group of $SL_{N+1}$, and
\item[(2)] the derived center of the enveloping algebra $U\fg$ is given by derived invariants of the adjoint action of $\fg$ on $U\fg$. That is to say $\Hoch^*(U\fg,U\fg) \simeq \clies(\fg, U\fg)$.
\end{enumerate}
\owen{And keep going just a little.}
%
%Our results pertains to (1) by taking $N \to \infty$, and to (2) through the derived {\em centralizer} of the map $\UU_n\sl_\infty \hookrightarrow \UU_n\fgl_\infty(A)$, where $A$ is a dg algebra. Note that the usual derived center is the centralizer of the identity map $\UU_n\fg \to \UU_n\fg$.
%
%We will review these notions and describe our results in detail Min the next section. When $n=1$,  \owen{state the classical thing here!}
%
%Our primary tool here is the Loday-Quillen-Tsygan (LQT) theorem which describes the large $N$ limit of the Lie algebra homology of $\fgl_N(A)$, for a dg associative algebra $A$. Since the enveloping $\EE_n$ algebra of a Lie algebra can be constructed as a locally constant factorization algebra made out of the shifted Chevalley-Eilenberg chains \cite{}, the relevance of the LQT theorem in this context  should be apparent. From this point of view, one can think of our main result as an $\EE_n$ generalization of the LQT theorem. We hope these ideas have exciting applications in quantum field theory and random matrix theory, as well as to other large $N$ limit phenomena.

@@

We apply our central result to important classes of Lie algebras: 
abelian, semisimple, and free.

@@

@@

In the case $k = 0$, we have the immediate corollary that
\[
\int_{S} \fZ_n(\fg) \simeq \fZ_0 \left( \int_S \UU_n \fg \right) ,
\]
which unwinds to 
\begin{equation}
\label{eqn: taut}
\clies(\fg \otimes \Omega^\bu(S), \cliels(\fg \otimes \Omega^\bu(S)) \simeq \End^\bullet(\cliels(\fg \otimes \Omega^\bu(S))),
\end{equation}
which looks rather tautological.
Nonetheless, by using other models for factorization homology,
one obtains less obvious identifications.

@@

\subsection{Cyclic homology and the large $N$ limits of $\UU_n \fgl_N$ and ${\frak Z}_n(\fgl_N)$}

In the case of the infinite series of Lie algebras, 
there are theorems that tell us about the limiting behavior of Lie algebra chains and cochains as $N$ goes to infinity.
These can be applied to our models to obtain striking results about factorization homology.

\def\Dih{{\rm Dih}}

Let $\Cyc_*(A)$ denote the {\em cyclic chains} of a dg associative algebra~$A$.
A key theorem --- one of the original motivations for cyclic homology --- is the Loday-Quillen-Tsygan (LQT) theorem \cite{},
which provides an explicit quasi-isomorphism
\begin{equation}
\label{eqn: LQT}
\cliels(\fgl_\infty(A)) \xto{\simeq} \Sym(\Cyc_*(A)[1])
\end{equation}
for any unital dg associative algebra~$A$.
A similar result \cite{} holds when we replace $\fgl_N$ by $\frak{so}_N$ or $\frak{sp}_N$, with
\[
\label{eqn: LQT}
\cliels({\frak{so}}_\infty(A)) \xto{\simeq} \Sym(\Dih_*(A)[1]),
\]
with $\Dih_*$ denoting the dihedral chains.
Note that one can use either $\frak{so}_N$ or $\frak{sp}_N$ on the left hand side.

\owen{Please verify my conditions, as I got a bit confused between different sources.}

Another key theorem, due to Jones \cite{} for cyclic chains and extend to dihedral chains by Ungheretti \cite{},
is that
\[
\Cyc_*(\Omega^*(M)) \simeq \Omega^*(LM/^h S^1)
\]
for a simply connected manifold $M$, 
where $LM/^h S^1$ denotes the homotopy quotient of the free loop space $LM$ under the canonical rotation action on loops,
and where we mean some model of cochains on the right hand side.
In other words, the right hand side is the $S^1$-equivariant cochains of the free loop space.
(This result holds, more generally, for a simply connected topological space of finite type, so long as one uses an appropriate model of cochains of the space, 
but we focus here on manifolds.)

From hereon we will work only with cyclic chains and $\fgl$,
but one obtains immediate analogs by replacing those with dihedral chains and $\frak{so}$ or~$\frak{sp}$.

As a first observation, let us see what this result implies about factorization homology of the enveloping $\EE_n$ algebra.

\begin{lmm}
For a closed $n$-dimensional manifold~$M$, 
\begin{align*}
\int_M \UU_n(\frak{gl}_\infty)
&\simeq \cliels(\Omega^*(M) \otimes \fgl_\infty) \\
&\simeq \Sym(\Cyc_*(\Omega^*(M))[1]) \\
&\simeq \Sym(\Omega^*(LM/^h S^1))[1])).
\end{align*}
\end{lmm}

Although this result is immediate from the theorems we just mentioned,
it is rather surprising to see the {\em cyclic loop space} $LM/^h S^1$ appears out of taking a large $N$ limit of factorization homology.
Note that this identification provides a commutative algebra structure on the factorization homology for $\fgl_\infty$, 
which is nowhere to be seen for~$\fgl_N$.

\begin{qtn}
Is there some local version of this relationship? That is, can we obtain a version of the LQT theorem at the level of $\EE_n$ algebras such that factorization homology recovers the quasi-isomorphism of this lemma?
\end{qtn}

This kind of result is especially interesting from the perspective of physics, 
as a kind of topological gauge/string duality.
In Section~\ref{sec: TFT} we explain how $\UU_n(\fgl_\infty)$ can be understood as the {\em quantum observables} of a gauge theory for the Lie algebra $\fgl_\infty$.
In this sense, the left hand side of the lemma $\int_M \UU_n(\frak{gl}_\infty)$ describes the global quantum observables over the manifold $M$.
On the right hand side, we see the cyclic loop space of $M$,
which arises naturally as the configuration space of a closed string theory with target~$M$.
Indeed, those familiar with closed string field theory will recognize the right hand side as related to the ``Fock space'' of such a theory.
\owen{Check me, Bri-guy.}

Now let's examine what our calculus results imply;
we simply apply quasi-isomorphism~\eqref{eqn: taut} in conjunction with the theorems.

\begin{lmm}
For a closed $n$-dimensional manifold~$M$, 
\[
\clies(\fgl_\infty \otimes \Omega^\bu(M), \cliels(\fgl_\infty \otimes \Omega^\bu(M)) \simeq \End^\bullet(\Sym(\Omega^*(LM/^h S^1))[1])))
\]
where the left hand side is the factorization homology of $\EE_n$ center ${\frak Z}_n(\fgl_\infty)$ and the right hand side is the $\EE_0$ center of the factorization homology of~$\UU_n \fgl_\infty$.
\end{lmm}

In physical terms, the right hand side encodes the algebra of operators on the Fock space of a closed string field theory.
Hence this statement provides a novel way to construct that algebra.
Even better, in Section~\ref{sec: TFT}, we identify the left hand side as the quantum observables of the topological BF theory for $\fgl_N$ on $M \times \RR$.
In this sense we have deepened the topological gauge/string duality just introduced.
